---
title:,"CHAP-5_BAYESIAN"
author: "Tawhid Tasdique"
date:,"2026-01-30"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pkg <- c("tidyverse","ggthemes", "utils")

lapply(X = pkg, FUN = library, character.only = T)

```

#Ex-5.1

##Ex-5.1(a)

```{r importing_data}

school1 <- read.table(file = "school1.dat", 
                      quote = "/", 
                      comment.char = "")

school2 <- read.table(file = "school2.dat", 
                      quote = "/", 
                      comment.char= "")

school3 <- read.table(file = "school3.dat", 
                      quote = "/", 
                      comment.char= "")

school1 <- school1 %>% 
  dplyr::select(V1) %>% 
  as.matrix() %>% 
  as.numeric()

school2 <- school2 %>% 
  dplyr::select(V1) %>% 
  as.matrix() %>% 
  as.numeric()

school3 <- school3 %>% 
  dplyr::select(V1) %>% 
  as.matrix() %>% 
  as.numeric()

```

## Theory

The kernel part derived in **HomeCopy-II** is non-standard or unfamiliar. So, we can't sample from it directly. Instead we will implement Monte-Carlo simulation to approximate $(\sigma^2 | y_1, ..., y_n)$ and $(\theta | \sigma^2, y_1, ..., y_n)$. 

First, we sample $\sigma^2$ from it's marginal posterior and then implement those values to sample $\theta$ from its full conditional distribution. 

We use the obtained Monte-Carlo samples to compute necessary means and $95$% confidence intervals. Since, posterior distributions are not purely theoretical quantities, we will compute (empirical) quantile-based confidence intervals from the produced samples. 

Note that, the theoretical mean of the full conditional distribution, $(\theta | \sigma^2, y_1, ..., y_n)$, is **NOT** the posterior mean for $\theta$. The "posterior mean" refers to the **marginal** posterior's mean. But that is an unrecognized distribution, at least for $\theta$ it is. Hence, we have to approximate it by averaging the relevant Monte-Carlo samples for each school.

```{r}

# Hyper-parameters
mu0 <- 5; sigma20 <- 4; k0 <- 1; v0 <- 2

# sample means
y1bar <- school1 %>% mean()
y2bar <- school2 %>% mean()
y3bar <- school3 %>% mean() 

# sample sizes
n1 <- school1 %>% length()
n2 <- school2 %>% length()
n3 <- school3 %>% length()

##############################################################

# sampling from 1st inverse-Gamma posterior
`sigma1^2|data` <- 1 / rgamma(
  n = 1000,
  shape = (v0 + n1) / 2,
  rate = (v0 * (sigma20^2) + var(school1)) / 2
)

# sampling from 2nd inverse-Gamma posterior
`sigma2^2|data` <- 1 / rgamma(
  n = 1000,
  shape = (v0 + n2) / 2,
  rate = (v0 * (sigma20^2) + var(school2)) / 2
)

# sampling from 3rd inverse-Gamma posterior
`sigma3^2|data` <- 1 / rgamma(
  n = 1000,
  shape = (v0 + n3) / 2,
  rate = (v0 * (sigma20^2) + var(school3)) / 2
)

##############################################################

# theoretical means from full conditional distribution of theta
mu1 <- weighted.mean(x = c(mu0, y1bar), # school 1 
                     w = c(k0, n1))
mu2 <- weighted.mean(x = c(mu0, y2bar), # school 2
                     w = c(k0, n2))
mu3 <- weighted.mean(x = c(mu0, y3bar), # school 3
                     w = c(k0, n3))

##############################################################

# sampling theta from 1st full conditional posterior
`theta|sigma1^2, data` <- sapply(
  X = `sigma1^2|data`, 
  FUN = function(sig){
    rnorm(n = 1, mean = mu1, sd = sig)
  }
)

# sampling theta from 2nd full conditional posterior
`theta|sigma2^2, data` <- sapply(
  X = `sigma2^2|data`, 
  FUN = function(sig){
    rnorm(n = 1, mean = mu2, sd = sig)
  }
)

# sampling theta from 2nd full conditional posterior
`theta|sigma3^2, data` <- sapply(
  X = `sigma3^2|data`, 
  FUN = function(sig){
    rnorm(n = 1, mean = mu3, sd = sig)
  }
)

##############################################################

# posterior means approximated for theta
post.mean.theta.1 <- mean(`theta|sigma1^2, data`)
post.mean.theta.2 <- mean(`theta|sigma2^2, data`)
post.mean.theta.3 <- mean(`theta|sigma3^2, data`)

# posterior means approximated for sigma
post.mean.sigma.1 <- mean(sqrt(`sigma1^2|data`))
post.mean.sigma.2 <- mean(sqrt(`sigma2^2|data`))
post.mean.sigma.3 <- mean(sqrt(`sigma3^2|data`))

##############################################################

# quantile-based 95% confidence interval for theta
post.ci.theta.1 <- quantile(x = `theta|sigma1^2, data`, 
                     probs = c(0.025, 0.975))
post.ci.theta.2 <- quantile(x = `theta|sigma2^2, data`, 
                     probs = c(0.025, 0.975))
post.ci.theta.3 <- quantile(x = `theta|sigma3^2, data`, 
                     probs = c(0.025, 0.975))

# quantile-based 95% confidence interval for sigma
post.ci.sigma.1 <- quantile(x = `sigma1^2|data`, 
                            probs = c(0.025, 0.975)) %>% sqrt() 
post.ci.sigma.2 <- quantile(x = `sigma2^2|data`, 
                            probs = c(0.025, 0.975)) %>% sqrt()
post.ci.sigma.3 <- quantile(x = `sigma3^2|data`, 
                            probs = c(0.025, 0.975)) %>% sqrt()

```

## Ex-5.1(b)



```{r}

theta1 <- `theta|sigma1^2, data`
theta2 <- `theta|sigma2^2, data`
theta3 <- `theta|sigma3^2, data`

abc <- mean(theta1 < theta2 & theta2 < theta3)
acb <- mean(theta1 < theta3 & theta3 < theta2)

bac <- mean(theta2 < theta1 & theta1 < theta3)
bca <- mean(theta2 < theta3 & theta3 < theta1)

cab <- mean(theta3 < theta1 & theta1 < theta2)
cba <- mean(theta3 < theta2 & theta2 < theta1)

```

## Ex-5.1(c)



```{r}



```












