---
title: "CHAP-5_BAYESIAN"
author: "Tawhid Tasdique"
date: "2026-01-30"
output:
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pkg <- c("tidyverse","ggthemes", "utils")

lapply(X = pkg, FUN = library, character.only = T)

set.seed(27)

```

# Ex-5.1

## Ex-5.1(a)

```{r importing_data}

school1 <- read.table(file = "school1.dat", 
                      quote = "/", 
                      comment.char = "")

school2 <- read.table(file = "school2.dat", 
                      quote = "/", 
                      comment.char= "")

school3 <- read.table(file = "school3.dat", 
                      quote = "/", 
                      comment.char= "")

school1 <- school1 %>% 
  dplyr::select(V1) %>% 
  as.matrix() %>% 
  as.numeric()

school2 <- school2 %>% 
  dplyr::select(V1) %>% 
  as.matrix() %>% 
  as.numeric()

school3 <- school3 %>% 
  dplyr::select(V1) %>% 
  as.matrix() %>% 
  as.numeric()

```

## Theory

The kernel part derived in **HomeCopy-II** is non-standard or unfamiliar. So, we can't sample from it directly. Instead we will implement Monte-Carlo simulation to approximate $(\theta | y_1, ..., y_n)$. 

First, we directly sample $\sigma^2$ from it's marginal posterior and then implement reuse those values to sample $\theta$ from its full conditional distribution i.e. $(\theta | y_1, ..., y_n)$. 

We use the obtained Monte-Carlo samples to compute necessary means and $95$% confidence intervals. Since, posterior distributions are not purely theoretical quantities, we will compute (empirical) quantile-based confidence intervals from the produced samples. 

Note that, the theoretical mean of the full conditional distribution, $(\theta | \sigma^2, y_1, ..., y_n)$, is **NOT** the posterior mean for $\theta$. The "posterior mean" refers to the **marginal** posterior's mean. But that is an unrecognized distribution, at least for $\theta$ it is. Hence, we have to approximate it by averaging the relevant Monte-Carlo samples for each school.

```{r}

# Hyper-parameters
mu0 <- 5; sigma20 <- 4; k0 <- 1; v0 <- 2

# sample means
y1bar <- school1 %>% mean()
y2bar <- school2 %>% mean()
y3bar <- school3 %>% mean() 

# sample sizes
n1 <- school1 %>% length()
n2 <- school2 %>% length()
n3 <- school3 %>% length()

# posterior parameters

kn1 <- k0 + n1
kn2 <- k0 + n2
kn3 <- k0 + n3

vn1 <- v0 + n1
vn2 <- v0 + n2
vn3 <- v0 + n3

##############################################################

# sampling from 1st inverse-Gamma posterior
`sigma1^2|data` <- 1 / rgamma(
  n = 1000,
  shape = vn1 / 2,
  rate = (vn1/2)*(
    v0*(sigma20^2) + 
    (n1-1)*var(school1) + 
    (k0*n1/kn1)*(y1bar-mu0)
  )
)

# sampling from 2nd inverse-Gamma posterior
`sigma2^2|data` <- 1 / rgamma(
  n = 1000,
  shape = vn2 / 2,
  rate = (vn2/2)*(
    v0*(sigma20^2) + 
    (n2-1)*var(school2) + 
    (k0*n2/kn2)*(y2bar-mu0)
  )
)

# sampling from 3rd inverse-Gamma posterior
`sigma3^2|data` <- 1 / rgamma(
  n = 1000,
  shape = vn3 / 2,
  rate = (vn3/2)*(
    v0*(sigma20^2) + 
    (n3-1)*var(school3) + 
    (k0*n3/kn3)*(y3bar-mu0)
  )
)

##############################################################

# theoretical means from full conditional distribution of theta
mu1 <- weighted.mean(x = c(mu0, y1bar), # school 1 
                     w = c(k0, n1))
mu2 <- weighted.mean(x = c(mu0, y2bar), # school 2
                     w = c(k0, n2))
mu3 <- weighted.mean(x = c(mu0, y3bar), # school 3
                     w = c(k0, n3))

##############################################################

# sampling theta from 1st full conditional posterior
`theta|sigma1^2, data` <- sapply(
  X = `sigma1^2|data`, 
  FUN = function(sig){
    rnorm(n = 1, mean = mu1, sd = sig)
  }
)

# sampling theta from 2nd full conditional posterior
`theta|sigma2^2, data` <- sapply(
  X = `sigma2^2|data`, 
  FUN = function(sig){
    rnorm(n = 1, mean = mu2, sd = sig)
  }
)

# sampling theta from 2nd full conditional posterior
`theta|sigma3^2, data` <- sapply(
  X = `sigma3^2|data`, 
  FUN = function(sig){
    rnorm(n = 1, mean = mu3, sd = sig)
  }
)

##############################################################

# posterior means approximated for theta
post.mean.theta.1 <- mean(`theta|sigma1^2, data`)
post.mean.theta.2 <- mean(`theta|sigma2^2, data`)
post.mean.theta.3 <- mean(`theta|sigma3^2, data`)

# posterior means approximated for sigma
post.mean.sigma.1 <- mean(sqrt(`sigma1^2|data`))
post.mean.sigma.2 <- mean(sqrt(`sigma2^2|data`))
post.mean.sigma.3 <- mean(sqrt(`sigma3^2|data`))

##############################################################

# quantile-based 95% confidence interval for theta
post.ci.theta.1 <- quantile(x = `theta|sigma1^2, data`, 
                     probs = c(0.025, 0.975))
post.ci.theta.2 <- quantile(x = `theta|sigma2^2, data`, 
                     probs = c(0.025, 0.975))
post.ci.theta.3 <- quantile(x = `theta|sigma3^2, data`, 
                     probs = c(0.025, 0.975))

# quantile-based 95% confidence interval for sigma
post.ci.sigma.1 <- quantile(x = `sigma1^2|data`, 
                            probs = c(0.025, 0.975)) %>% sqrt() 
post.ci.sigma.2 <- quantile(x = `sigma2^2|data`, 
                            probs = c(0.025, 0.975)) %>% sqrt()
post.ci.sigma.3 <- quantile(x = `sigma3^2|data`, 
                            probs = c(0.025, 0.975)) %>% sqrt()

```

## Ex-5.1(b)



```{r}

theta1 <- `theta|sigma1^2, data`
theta2 <- `theta|sigma2^2, data`
theta3 <- `theta|sigma3^2, data`

(abc <- mean(theta1 < theta2 & theta2 < theta3))

(acb <- mean(theta1 < theta3 & theta3 < theta2))

(bac <- mean(theta2 < theta1 & theta1 < theta3))

(bca <- mean(theta2 < theta3 & theta3 < theta1))

(cab <- mean(theta3 < theta1 & theta1 < theta2))

(cba <- mean(theta3 < theta2 & theta2 < theta1))

```

## Ex-5.1(c)

Now, we need to simulate data from the posterior predictive distributions. The form of the posterior predictive distribution is too complicated and time-consuming to derive if not impossible. So, we're going to employ Monte-Carlo simulation to approximate it instead.

```{r}

# sampling from school1's posterior predictive distribution
y1pred <- sapply(
  X = 1:length(`sigma1^2|data`), 
  FUN = function(i){
    rnorm(n = 1, 
          mean = `theta|sigma1^2, data`[i], 
          sd = `sigma1^2|data`[i])
  }
)

# sampling from school2's posterior predictive distribution
y2pred <- sapply(
  X = 1:length(`sigma2^2|data`), 
  FUN = function(i){
    rnorm(n = 1, 
          mean = `theta|sigma2^2, data`[i], 
          sd = `sigma2^2|data`[i])
  }
)

# sampling from school3's posterior predictive distribution
y3pred <- sapply(
  X = 1:length(`sigma3^2|data`), 
  FUN = function(i){
    rnorm(n = 1, 
          mean = `theta|sigma3^2, data`[i], 
          sd = `sigma3^2|data`[i])
  }
)

(abc <- mean(y1pred < y2pred & y2pred < y3pred))

(acb <- mean(y1pred < y3pred & y3pred < y2pred))

(bac <- mean(y2pred < y1pred & y1pred < y3pred))

(bca <- mean(y2pred < y3pred & y3pred < y1pred))

(cab <- mean(y3pred < y1pred & y1pred < y2pred))

(cba <- mean(y3pred < y2pred & y2pred < y1pred))

```

## Ex-5.1(d)



```{r}

(ans1 <- mean(theta1 > theta2 & theta1 > theta3))

(ans2 <- mean(y1pred > y2pred & y1pred > y3pred))

```

# Ex-5.2

What is sensitivity analysis? Explain.

```{r}

# present sample sizes
nA <- nB <- 16 

# hyper-parameters
mu0 <- 75; sigma0 <- 10

# sample means
ybarA <- 75.2; ybarB <- 77.5

# sample standard deviation
SA <- 7.3; sB <- 7.3

# prior sample sizes
k0 <- v0 <- 2^c(0:5)
# what is the purpose of v0??

# Since it is unknown, we will use the sample variance (`s^2`) as an estimate of the true sampling variability or, population variance (`sigma^2`).


```

# Ex-5.3 

I've already derived all the distributions in **Assignment-4** and **HomeCopy-3**. **Assignment-4** can be found on **HomeCopy-1** and a part of it can also be found on **HomeCopy-3**. The marginal posterior distribution for $\theta$ is a non-standard distribution. So, it will be approximated via Monte-Carlo sampling. Then I will compare it with closely-spaced densities computed by a user-defined PDF (probability density function).

```{r}

set.seed(2)

# kernel part of marginal posterior of theta
# this function needs to be modified to abide by total law of probability
# I'm going to try discrete approximation and see what happens
mp.theta <- function(theta, k0 = 20, mu0 = 5.64, v0 = 7, sigma20 = 27.43, y){
  
  # data-related values
  n <- length(y)
  ybar <- mean(y)
  
  # posterior parameters
  kn <- k0 + n
  vn <- v0 + n
  mun <- (k0*mu0 + n*ybar)/kn
  sigma2n <- (v0*sigma20 + (n-1)*var(y) + (k0*n/kn)*(ybar - mu0)^2)/vn
  
  # numerator of pdf
  numerator <- (1/sqrt(2*pi/k0))*((vn*sigma2n/2)^(vn/2))*gamma((vn+1)/2)
  
  # denominator of pdf
  denominator <- gamma(vn/2)*((kn*(theta-mun)^2 + 
                                 vn*sigma2n)/2)^((vn+1)/2)
  
  # kernel part of the pdf 
  `P(theta|y)` <- numerator/denominator
  
}

# hyper-parameter values (these were arbitrarily chosen)
k0 <- 20
v0 <- 7
mu0 <- 5.64
sigma20 <- 27.43

# sample data
y <- rnorm(n = 50, mean = 9, sd = 7) 
n <- length(y)
ybar <- mean(y)

# updated parameters (based on hyper-parameters and data)
kn <- k0 + n
vn <- v0 + n
mun <- (k0*mu0 + n*ybar)/kn
sigma2n <- (v0*sigma20 + (n-1)*var(y) + (k0*n/kn)*(ybar - mu0)^2)/vn

# sample of variances from the inverse-gamma marginal posterior
`sigma^2|y` <- 1/rgamma(n = 1000, 
                        shape = vn/2, 
                        rate = vn*sigma2n/2)

# Monte-Carlo Approximation for marginal posterior of theta
`theta|sigma^2,y` <- sapply(
  X = `sigma^2|y`, 
  FUN = function(variance){
    rnorm(n = 1, 
          mean = mun, 
          sd = sqrt(variance))
  }
)

# Monte-Carlo Approximated Density (visuals)
tibble(theta.post = `theta|sigma^2,y`) %>%
  ggplot(aes(x = theta.post)) +
  # 1st layer
  geom_histogram(aes(y = after_stat(density)),
                 color = "darkblue",
                 fill = "lightblue") +
  # 2nd layer
  geom_line(data = tibble(
    theta.values = (-20):40,
    theta.density = sapply(
      X = theta.values,
      FUN = function(val) {
        # set.seed(27) # to reproduce the same sample
        mp.theta(theta = val, y = y)
      }
    )
  ), 
  aes(x = theta.values, y = theta.density), 
  color = "red", 
  linewidth = 1)

```

